Investigating soundscapes perception through acoustic scenes simulation.

### 2017-09-01 lafaySimulatedScenePerception.tex
improve introduction

  \begin{keyword}
  %% keywords here, in the form: keyword \sep keyword
 cognitive psychology \sep soundscape assessment
  %% PACS codes here, in the form: \PACS code \sep code

@@

One of the main goal of soundscape studies is to identify which components of the soundscape influence human perception \cite{aletta2016soundscape}. For example, in order to study the link between an environment (urban sound scene) and the induced human sensation (calmness), one mean is to objectify the mental representation of a specific sound scene, a calm urban sound scene. Several means have been considered in the literature to perform this task.

@@

First, a subject can be asked to assess a given sound following a given perceptual scale (\eg~calmness, pleasantness) \cite{axelsson2005soundscape,davies2013perception,cain2013development}. The amount of information that can be gathered strongly depends on the very nature of the available stimuli, be they recorded sound scenes as part of a within laboratory experiment, or \emph{in situ}. That being said, those recorded stimuli  can be analysed, what allows the experimenter to gather a coarse
grain physical description of them.

@@

For several reasons that are detailed in this paper, the use of a soundscape simulator such as the one proposed in this article can lead to interesting outcomes as it allows the experimenter to separately study  the influence of the sound sources that compose a sound environment of interest. With such a material, not only the type of sound sources is indeed available for study, but also the exact level and audio waveform for each source together with the structural characteristics of the scene, that is the temporal distribution of the events.

@@

@@

To the best of our knowledge, only Bruce~\cite{bruce2009development,bruce2014effects} considered the use of a simulator to question soundscape perception. They propose a tool that allows the user to modify a given soundscape by adding or removing specific sound sources, by changing the acoustic level of the sources as well as their spatial location. The authors shows that the addition or removal of the sources globally follows more social or semantic matters than their acoustical characteristics. A lack of diversity is nonetheless mentioned by the authors that limit the strenght of the outcomes given in this study.

@@

\emph{Simscene} is an online digital audio tool whose first version has been developed as part of the HOULE project\footnote{\emph{Projet HOULE} : \url{houle.ircam.fr}}. It has been designed to run on the popular web browsers \emph{Chrome} and \emph{Firefox}. It is fully written in javascript using the \emph{angular.js} library\footnote{\emph{angular.js} : \url{angularjs.org}} and the \emph{web-audio} standard that allows the manipulation of digita audio data within the browser\footnote{\emph{web-audio}: \url{www.w3.org/TR/webaudio}}. The interface for selecting the sound sources (\cf~Section~\ref{sec:simscene_ssf}) uses the popular \emph{D3.js} \cite{d32011} visualization library.

@@

The literature usually distinguishes between:
\begin{itemize}
\item {a sound event}: an isolated sound, sparsely distributed whose acoustical characteristics may change with respect to time
\item {a sound texture}: an isolated sound of long duration whose acoustical characteristics are stable with respect to time \cite{saint1995classification}.
\end{itemize}

The simulation tool \emph{Simscene} follows this distinction and considers two distinct sound databases: one with the classes of events only and the other with classes of textures only. These two types also have specific scheduling procedures during the simulation process.

@@

@@

According to the Auditory Scene Analysis (ASA) theory, the brain processes separately events emerging from different sources of a sound environment \cite{bregman1994auditory}. Several auditory streams are created, one for each sequence of events emitted by the same source \cite{carlyon2004brain}. In this case, the source is a perceptual entity as events that are not isolated are grouped into a stream.

@@

The sound classes are organized hierarchically (\cf~Figure~\ref{fig:orgDb}) according to a structure that is close to the one of the vertical axis of the categorical organization proposed by E. Rosch \cite{rosch1978cognition}. The lower the level of abstraction, the more precise the description of the class and the more perceptually similar the sound sources is. For classes with high level of abstraction that have sub classes, their collection of samples is the union of the collections of the sub classes.

@@

By a volountary design choice, the simulation tool does not allow the user to interact and control directly a specific sample. Interaction is done at the track level, a track being a sequence of samples. Several parameters are available to the subject to control the track:

@@

\begin{itemize}
\item \emph{level} ($dB$): the levels are  randomly sampled according to a normal distribution parametrized by the subject in terms of mean value and variance;
\item \emph{inter-onset spacing} (second): for the event's track only, the spacings are randomly sampled according to a normal distribution parametrized by the subject in terms of mean value and variance;
\item \emph{start and end time} (second): the subject set the first and last occurrence of a sample of the selected class.
\end{itemize}

@@

\begin{itemize}
\item \emph{event's fades} (seconds): for the event's track only, the subject can set a fade in / fade out duration applied to each samples;
\item \emph{global fades} (seconds): the subject set the duration of the fade in / fade out at the beginning and the end of the track.
\end{itemize}

@@

@@

\begin{figure}
        \subfloat[]
        {\includegraphics[width=.8\linewidth]{gfx/ch_4/SSF}\label{fig:ssf}} \par
        \subfloat[]
        {\includegraphics[width=\linewidth]{gfx/ch_4/simscene}\label{fig:simscene}}
       \caption{Graphical interfaces for the selection of sound classes (a) and their sequencing (b) of the \emph{simScene} tool.}
\end{figure}

@@

A set of features is attached to each sound scene. The analysis is done on the basis of these features. A summary of these features (and the corresponding acronyms) is presented in Table ~\ref{tab:acronyme}. In order be consistent with the evaluation of experiment 1.b, features are not computed in the whole duration of the sequences but only on their 30-second reduced version used as stimuli for experiment 1.b, see Section \ref{sec:xp1b_plan}.
For each sound scene, three types of features are considered :
\begin{itemize}
\item \emph{perceptual}: the perceived pleasantness of the composed scene, assessed on a 7-point Likert scale. $\mathcal{A}_{scene}$ is the average pleasantness for one scene, as the average of the marks of all the participants. $\mathcal{A}_{subject}$ is the average pleasantness for each participant, computed as the average of all the score given by each participant. Given the low number of participants in Exp. 1.b, we choose not to normalize the pleasantness scores.
\item \emph{semantic}: a boolean vector $S = (x_1, x_2, \ldots, x_n)$ that indicates the class of sounds involved in the scene. Each boolean $x$ of this vector corresponds to a specific class of sounds: $x = 1$ if the class is present in the scene, and $x = 0$ otherwise. The vector dimension ($n$) depends on the level of abstraction that is considered for analysis. For instance, for the abstraction level 1 that includes 44 classes of sounds, $n = 44$.
\item \emph{structural}: features related to \emph{sound level}. To figure out the sound level, we take the $L_{Aeq}$ feature as a conceptual reference. In our case, it is a scalar computed from the  signal (in Volt and not in Pascal), and converted in decibels, with a reference of 1 Volt (full scale). The level is obtained by computing the quadratic mean of the signal every second and averaging the results over the total duration of the scene. An A-filtering module processes the data before the quadratic means are computed. We note $L$, $L(E)$ and $L(T)$ the computed levels by respectively considering the whole set of samples, only the set of event samples and only the set of texture samples.
\end{itemize}

@@

\begin{itemize}
\item \emph{checking the pleasantness of the simulated sound scene}: to check that the emotional difference produced by the i- and ni-scenes is also present in terms of perceived pleasantness, we look at the existence of differences between the average $\mathcal{A}_{scene}$ and the average $\mathcal{A}_{subject}$ for each type of scenes (i or ni). The significance is evaluated by a Student test on two independent populations for $\mathcal{A}_{scene}$ and by a Student test on two paired populations for $\mathcal{A}_{subject}$;
\jfp{ce qu'est Ascene dans ce paragraphe n'est pas clair :
A scene est défini comme la moyenne des scores d'une scène donnée pour tous les participants.
le test, lui, porte sur la moyenne des Ascene pour les deux catégories
meme chose pour les sujets}
\item \emph{analysis of the structural features}: in order to evaluate whether the affective difference between  the i- and ni-scenes has a significative influence  on the structural features of the scenes \ie~ if significative differences exist between structural features and perceived pleasantness. The significance is measured with a Student test on two populations \jfp{je ne comprends pas bien ce test}
\item \emph{study of the influence of the structural features on the perceived pleasantness}: in order to evaluate a possible impact of the structural features on the perceived pleasantness, linear correlations between these features and $\mathcal{A}_{scene}$ are considered. The Pearson correlation coefficient is used for that purpose, on the basis of the standard methodology used in multidimensional studies;
\item \emph{analysis of the semantic features}: in order to determine if the affective difference between  the i- and ni-scenes has an influence on the scenes composition in terms of sound sources, or if specific classes are more frequently used to simulate a given environment (i or ni), a V-test is considered. Testing is done at every level of abstraction, and separately for texture and event classes. For each class $j$, and each environment type $k$ ($k=\lbrace i,ni\rbrace$), the value $V_{jk}$ is computed as follows:

@@

where $c$ is the number of available classes, $c_k$ the number of classes for a given type of environment $k$, $c_j$ the number of  $j$ classes, and $c_{jk}$ the number of $j$ classes used for an environment $k$. \jfp{il faudrait donner une référence pour ce test : bouquin Lebart, Morineau, Piron, STATISTIQUE EXPLORATOIRE MULTIDIMENSIONNELLE
peut être donner les bornes de variations de Vjk pour aider le lecteur à comprendre cet indicateur} \ml{y pas une ref en Anglais ?} The V-test tests the null hypothesis that the ratio $\frac{c_{jk}}{c}$ is not significatively different from the ratio $\frac{c_{jk}}{c_k}$. For a given environment $k$, and class $j$, if the hypothesis is rejected, the class $j$ is said to be typical with respect to environment $k$. The typical classes are called \textbf{sound markers};\jfp{on ne comprend pas bien l'hypothèse alternative du test. VOIR BOUQUIN LEBART page 184 ou bien article de Morineau sur le sujet Vjk est une estimation approchée de la valeur test}

\item \emph{Study of the representation spaces induced by the semantic features}: in order to determine if a representation only based on the presence / absence of sound sources allows us to distinguish between the two types of environment, we investigate the space induced by the semantic features $S$. $S$ is a boolean vector and the distance between two scenes is computed with the Hamming distance. Considering two $n$-dimension vectors $S_1=(x_{1,1},x_{1,2},\ldots,x_{1,n})$, where $S_2=(x_{2,1},x_{2,2},\ldots,x_{2,n})$ is of cardinality $n$, with $x \in\lbrace 0,1\rbrace$, the Hamming distance $d_{ham}$ measures the proportion of coordinates that differ between two vectors, ans is defined as follows:

@@

where $\bigoplus$ is the \emph{exclusive-or}. The more the composition of the two scenes are similar, the closer they are in this space. The use of Hamming distance allows us to take into account equally the presence and absence of classes. For measuring the intrinsic ability of the space to split the i- and ni-scene, we use a ranking metric named the precision at rank $k$ ($P@k$) . The $P@k$ computes the precision obtained after the $k$ closest items of a given seed item have been found. Formally, for each $s_i$ scene (considered as seed), we compute the ratio between the number of $s_j$ scenes taken among the $k$ nearest neighbors of $s_i$ and sharing the same label than $s_i$ versus $k$. the $P@k$ is then the average of this ratio for all the items considered as search seeds;

@@

Except for the V-test, the statistical tests of significance are done with a critical threshold of $\alpha=0.05$. For the V-test, considering that a great number of classes is tested, a Bonferroni correction is applied. For the $p$-value, if $p\geq0.05$, the value is reported. If $0.01\leq p<0.05$, we only report $p<0.05$, otherwise we report $p<0.01$.

@@

\subsubsection*{Checking of the pleasantness of the synthesized scenes}

@@

\begin{table*}[t]
 \setlength{\tabcolsep}{0.2pt}
 \centering
  {\renewcommand{\arraystretch}{0.9}
\begin{tabular}{c c c c}
Abstraction        & \multicolumn{2}{c}{Event sound markers} \\
level & i-scenes & ni-scenes \\
\hline
0  &                          & construction work (3.78)  \\
\hline
  & church bell  (4.5)             & car horn  (3.9) \\
1 & bicycle bell  (4.3)      & siren (3.9)\\
  & animal (4.2)              &       \\
   \hline
  & bird        (4.8)       & car horn  (4.0)\\
2 & church bell  (4.4)             & siren (4.0)\\
  & bicycle bell     (4.2)             &       \\
   \hline
  & bird song (4.8)        & klaxon  (4.1)\\
3 & church bell   (4.3)            & siren (4.0)\\
  & bicycle bell      (4.2)   &       \\
  & foot steps  (3.6)      &  \\
  &                           & \\
  & \multicolumn{2}{c}{Texture sound markers}      \\
  & i-scenes & ni-scenes \\
\hline
0 &     courtyard / park (4.1) &  construction (3.9)  \\
\hline
1 &     park (3.65)          &  crossroad (3.6)  \\
  &                          &  construction vehicule (3.3)  \\
\hline
2 &     park (3.64)          &  crossroad (3.56)  \\
\hline
\end{tabular}
}
\vspace{0.5mm}
\caption{Event and texture classes identified as sound markers. In each cell, markers are ranked in the decreasing order of V-test value.\jfp{a partir de quelle valeur de V-test le marker est présent dans ce tableau ? mettre la p-value , non ? }}
\label{tab:markers}
\end{table*}

@@

The subjects evaluate the 144 scenes. The evaluation is done on a 11-point bipolar semantic scale ranging from -5 (non-ideal / very unpleasant) to +5 (ideal / very pleasant). Before scoring a scene, subjects must listen to the first 20 seconds. After scoring, they are free to move on to the next scene. \jfp{on se demande pourquoi on va de -5 a +5 alors qu'avant on allait de -3 à +3}

@@

The aim here is to validate the hypothesis that the removal of i-markers and ni-markers have an impact on the perceived pleasantness. To do so, we perform a an Analysis of Variance. We consider $\mathcal{A}_{subject}$ as a dependent variable \jfp{je crois que la défintion de la variable dépendante, Asubject, n'est pas la bonne. A subject a a été définit avant comme "Asubject is
the average pleasantness for each participant". l'ANOVA à deux facteurs à mesures répétées ne porte pas sur la valeur moyenne mais sur la valeur de l'évaluation pour une scene donnée. il faut donc définir une notation appropriée}, and as independent variables, the type of environment (i/ni), and the presence/absence of markers (km/rm). As each subject evaluated the whole set of scenes, a two factors repeated measure ANOVA is used to check the validity of the hypothesis that there exist a significant difference of perceived pleasantness between km-scenes and rm-scenes. The two independent variables are considered as \emph{within-subject} factors. The factors being of only two levels each (type: i/ni,  marker: km/rm), the sphericity hypothesis does not need to be checked. \emph{Post hoc} analyses are done using the Tukey-Kramer procedure.

@@

The \emph{post hoc} analysis exhibits significative differences between all the groups of observations, notably between the i/km-scenes ($p<0.05$) and the i/rm-scenes ($p<0.01$). \jfp{je ne comprends pas les 2 p-value. il ne peut y a voir qu'une seule p-value lorsqu'on compare deux conditions expérimentales
}

>>>>

  \begin{keyword}
  %% keywords here, in the form: keyword \sep keyword
 cognitive psychology \sep soundscape perception \ soundscape simulator
  %% PACS codes here, in the form: \PACS code \sep code

@@

One of the main goal of soundscape studies is to identify which components of the soundscape influence human perception \cite{aletta2016soundscape}. For example, in order to study the link between an environment (urban sound scene) and the induced human sensation (calmness), one mean is to objectify the mental representation of a specific sound scene (a calm urban sound scene). Several means have been considered in the literature to perform this task.

@@

First, a subject can be asked to assess a given sound following a given perceptual scale (\eg~calmness, pleasantness) \cite{axelsson2005soundscape,davies2013perception,cain2013development}. The amount of information that can be gathered strongly depends on the very nature of the available stimuli, be they recorded sound scenes as part of a within laboratory experiment, or \emph{in situ}. That being said, those recorded stimuli  can be analyzed, what allows the experimenter to gather a coarse
grain physical description of them.

@@

@@

For several reasons that are detailed in this paper, the use of a soundscape simulator such as the one proposed in this article can lead to interesting outcomes as it allows the experimenter to separately study  the influence of the sound sources that compose a sound environment of interest. With such a material, not only the type of sound sources is indeed available for study, but also the exact level and audio waveform for each source together with the structural characteristics of the scene, that is the temporal distribution of the sound events.

@@

To the best of our knowledge, only Bruce~\cite{bruce2009development,bruce2014effects} considered the use of a simulator to question soundscape perception. They propose a tool that allows the user to modify a given soundscape by adding or removing specific sound sources, by changing the acoustic level of the sources as well as their spatial location. The authors shows that the addition or removal of the sources globally follows more social or semantic matters than their acoustical characteristics. A lack of diversity in terms of sound sources is nonetheless mentioned by the authors that limit the strength of the outcomes given in this study.

@@

\emph{Simscene} is an online digital audio tool whose first version has been developed as part of the HOULE project\footnote{\emph{Projet HOULE} : \url{houle.ircam.fr}}. It has been designed to run on the popular web browsers \emph{Chrome} and \emph{Firefox}. It is fully written in javascript using the \emph{angular.js} library\footnote{\emph{angular.js} : \url{angularjs.org}} and the \emph{web-audio} standard that allows the manipulation of digital audio data within the browser\footnote{\emph{web-audio}: \url{www.w3.org/TR/webaudio}}. The interface for selecting the sound sources (\cf~Section~\ref{sec:simscene_ssf}) uses the popular \emph{D3.js} \cite{d32011} visualization library.

@@

The literature usually distinguishes between:

\begin{itemize}
\item {a sound event}: an isolated sound, sparsely distributed whose acoustical characteristics may change with respect to time
\item {a sound texture}: an isolated sound of long duration whose acoustical characteristics are stable with respect to time \cite{saint1995classification}.
\end{itemize}

@@

The simulation tool \emph{Simscene} follows this distinction and considers two distinct sound databases: one with the classes of events only and the other with the classes of textures only. These two types of sound classes also have specific scheduling procedures during the simulation process.

@@

According to the Auditory Scene Analysis (ASA) theory, the brain processes separately events emerging from the different sources of a sound environment \cite{bregman1994auditory}. Several auditory streams are created, one for each sequence of events emitted by the same source \cite{carlyon2004brain}. In this case, the source is a perceptual entity as events that are not isolated are grouped into a stream.

@@

The sound classes are organized hierarchically (\cf~Figure~\ref{fig:orgDb}) according to a structure that is close to the one of the vertical axis of the categorical organization proposed by E. Rosch \cite{rosch1978cognition}. The lower the level of abstraction, the more precise the description of the class and the more perceptually similar the sound sources. For classes with high level of abstraction that have sub classes, their collection of samples is the union of the collections of the sub classes.

@@

By a voluntary design choice, the simulation tool does not allow the user to interact and control directly a specific sample. Interaction is done at the track level, a track being a sequence of samples. Several parameters are available to the subject to control the track:

@@

\begin{itemize}
\item \emph{sound level} ($dB$): for each sample, the sound levels are randomly sampled according to a normal distribution parametrized by the subject in terms of mean value and variance;
\item \emph{inter-onset spacing} (second): for the event's track only, and for each sound event sample, the inter-onset spacings are randomly sampled according to a normal distribution parametrized by the subject in terms of mean value and variance;
\item \emph{start and end time} (second): the subject set the first and last occurrence of a sample of the selected class.
\end{itemize}

@@

\begin{itemize}
\item \emph{event's fades} (seconds): for the event's track only, the subject can set a fade in / fade out duration applied to each sample;
\item \emph{global fades} (seconds): the subject can set global fade in / fade out durations applied to the entire track.
\end{itemize}

@@

\begin{figure}
		\centering
        \subfloat[]
        {\includegraphics[width=.8\linewidth]{gfx/ch_4/SSF}\label{fig:ssf}} \par
        \subfloat[]
        {\includegraphics[width=\linewidth]{gfx/ch_4/simscene}\label{fig:simscene}}
       \caption{Graphical interfaces for the selection of sound classes (a) and their sequencing (b) of the \emph{simScene} tool.}
\end{figure}

@@

@@

A set of features is attached to each sound scene. The analysis is done on the basis of these features. A summary of these features (and the corresponding acronyms) is presented in Table ~\ref{tab:acronyme}. In order be consistent with the evaluation of experiment 1.b, features are not computed in the whole duration of the sequences but only on their 30-second reduced version used as stimuli for experiment 1.b, see Section \ref{sec:xp1b_plan}.

For each sound scene, three types of features are considered :

\begin{itemize}
\item \emph{perceptual}: the perceived pleasantness of the composed scene, assessed on a 7-point Likert scale. $\mathcal{A}_{scene}$ is the average pleasantness for one scene, as the average of the marks of all the participants. $\mathcal{A}_{subject}$ is the average pleasantness for each participant, computed as the average of all the score given by each participant. Given the low number of participants in Exp. 1.b, we choose not to normalize the pleasantness scores.
\item \emph{semantic}: a boolean vector $S = (x_1, x_2, \ldots, x_n)$ that indicates the class of sounds involved in the scene. Each boolean $x$ of this vector corresponds to a specific class of sounds: $x = 1$ if the class is present in the scene, and $x = 0$ otherwise. The vector dimension ($n$) depends on the level of abstraction that is considered for analysis. For instance, for the abstraction level 1 that includes 44 classes of sounds, $n = 44$.
\item \emph{structural}: features related to \emph{sound level}. To figure out the sound level, we take the $L_{Aeq}$ feature as a conceptual reference. In our case, it is a scalar computed from the  signal (in Volt and not in Pascal), and converted in decibels, with a reference of 1 Volt (full scale). The level is obtained by computing the quadratic mean of the signal every second and averaging the results over the total duration of the scene. An A-filtering module processes the data before the quadratic means are computed. We note $L$, $L(E)$ and $L(T)$ the computed levels by respectively considering the whole set of samples, only the set of event samples and only the set of texture samples.
\end{itemize}

@@

\begin{itemize}
\item \emph{analysis of the perceived pleasantness}: the goal is to evaluate whether the perceived pleasantness differ between the i- and ni-scenes. We look at the existence of differences between the average $\mathcal{A}_{scene}$ and the average $\mathcal{A}_{subject}$ for each type of scenes (i or ni). The significance is evaluated by a two-sample Student test for $\mathcal{A}_{scene}$ and by a paired-sample Student test for $\mathcal{A}_{subject}$;

%\jfp{ce qu'est Ascene dans ce paragraphe n'est pas clair :A scene est défini comme la moyenne des scores d'une scène donnée pour tous les participants. le test, lui, porte sur la moyenne des Ascene pour les deux catégories meme chose pour les sujets}

\item \emph{analysis of the structural features}: the goal is to evaluate whether the structural features differ between the i- and ni-scenes. The significance is measured with a two-sample Student test;

\item \emph{influence of the structural features on the perceived pleasantness}: the goal is to evaluate if structural features impact the perceived pleasantness. We look at linear correlations between these features and $\mathcal{A}_{scene}$. The Pearson correlation coefficient is used for that purpose;

\item \emph{analysis of the semantic features}: the goal is to evaluate if specific classes are more frequently used to simulate a given type of environment (i or ni). A V-test is considered \citep{lebart2000statistique} \ml{y pas une ref en Anglais ?}. With $c$ being the total number of classes used for both types of environment, $c_k$ the number of classes used for a given type of scene $k$ ($k=\lbrace i,ni\rbrace$), $c_j$ the number of time a class $j$ has been used, and $c_{jk}$ the number of time a class $j$ has been used for a given type of scene $k$, the V-test tests the null hypothesis that the ratio $\frac{c_{jk}}{c}$ is not significatively different from the ratio $\frac{c_{jk}}{c_k}$. For each class $j$, and each environment type $k$, an approximation of the statistic value $V_{jk}$ is computed as follows:

@@

If the hypothesis is rejected, the class $j$ is said to be typical with respect to the type of scene $k$. The typical classes are called \textbf{sound markers}. Testing is done for each class, at every level of abstraction, and separately for texture and event classes;

%\jfp{il faudrait donner une référence pour ce test : bouquin Lebart, Morineau, Piron, STATISTIQUE EXPLORATOIRE MULTIDIMENSIONNELLE peut être donner les bornes de variations de Vjk pour aider le lecteur à comprendre cet indicateur}  \jfp{on ne comprend pas bien l'hypothèse alternative du test. VOIR BOUQUIN LEBART page 184 ou bien article de Morineau sur le sujet Vjk est une estimation approchée de la valeur test}

\item \emph{representation space induced by the semantic features}: the goal is to determine if a representation space of the scenes solely based on the presence / absence of sound sources allows us to distinguish between the two types of scene. Considering $S_i$ the semantic feature of the scene $i$, we compute the distances between all the vectors $S_i$. A Hamming distance is used. Considering two $n$-dimension vectors $S_1=(x_{1,1},x_{1,2},\ldots,x_{1,n})$ and $S_2=(x_{2,1},x_{2,2},\ldots,x_{2,n})$, with $x \in\lbrace 0,1\rbrace$, the Hamming distance $d_{ham}$ measures the proportion of coordinates that differ between the two vectors. It is defined as follows:

@@

where $\bigoplus$ is the \emph{exclusive-or}. Two scenes having similar source compositions will be close in such space. The use of Hamming distance allows us to take into account equally the presence and absence of classes. For measuring the intrinsic ability of the space to split the i- and ni-scene, we use a ranking metric named the precision at rank $k$ ($P@k$). The $P@k$ computes the precision obtained after the $k$ closest items of a given seed item have been found. Formally, for each $s_i$ scene (considered as seed), we compute the ratio between the number of $s_j$ scenes taken among the $k$ nearest neighbors of $s_i$ and sharing the same label than $s_i$ versus $k$. the $P@k$ is then the average of this ratio for all the items considered as search seeds;

@@

The statistical tests of significance are done with a critical threshold of $\alpha=0.05$. For the V-test, considering that a great number of classes is tested, a Bonferroni correction is applied. For the $p$-value, if $p\geq0.05$, the value is reported. If $0.01\leq p<0.05$, we only report $p<0.05$, otherwise we report $p<0.01$.

@@

\subsubsection*{Analysis of the perceived pleasantness}

@@

\begin{table*}[t]
 \setlength{\tabcolsep}{0.2pt}
 \centering
  {\renewcommand{\arraystretch}{0.9}
\begin{tabular}{c c c c}
Abstraction        & \multicolumn{2}{c}{Event sound markers} \\
level & i-scenes & ni-scenes \\
\hline
0  &                          & construction work (3.78)  \\
\hline
  & church bell  (4.5)             & car horn  (3.9) \\
1 & bicycle bell  (4.3)      & siren (3.9)\\
  & animal (4.2)              &       \\
   \hline
  & bird        (4.8)       & car horn  (4.0)\\
2 & church bell  (4.4)             & siren (4.0)\\
  & bicycle bell     (4.2)             &       \\
   \hline
  & bird song (4.8)        & klaxon  (4.1)\\
3 & church bell   (4.3)            & siren (4.0)\\
  & bicycle bell      (4.2)   &       \\
  & foot steps  (3.6)      &  \\
  &                           & \\
  & \multicolumn{2}{c}{Texture sound markers}      \\
  & i-scenes & ni-scenes \\
\hline
0 &     courtyard / park (4.1) &  construction (3.9)  \\
\hline
1 &     park (3.65)          &  crossroad (3.6)  \\
  &                          &  construction vehicule (3.3)  \\
\hline
2 &     park (3.64)          &  crossroad (3.56)  \\
\hline
\end{tabular}
}
\vspace{0.5mm}
\caption{Event and texture classes identified as sound markers. In each cell, markers are ranked in the decreasing order of V-test value. $p\leq0.01$ for all sound markers.}
\label{tab:markers}
\end{table*}

@@

The subjects evaluate the 144 scenes. The evaluation is done on a 11-point bipolar semantic scale ranging from -5 (non-ideal / very unpleasant) to +5 (ideal / very pleasant). Before scoring a scene, subjects must listen to the first 20 seconds. After scoring, they are free to move on to the next scene.

%\jfp{on se demande pourquoi on va de -5 a +5 alors qu'avant on allait de -3 à +3}

@@

The aim here is to validate the hypothesis that the removal of i-markers and ni-markers have an impact on the perceived pleasantness. To do so, we perform an Analysis of Variance. We consider $\mathcal{A}_{subject}$ as a dependent variable, and as independent variables, the type of environment (i/ni), and the presence/absence of markers (km/rm). As each subject evaluated the whole set of scenes, a two factors repeated measure ANOVA is used to check the validity of the hypothesis that there exist a significant difference of perceived pleasantness between km-scenes and rm-scenes. The two independent variables are considered as \emph{within-subject} factors. The factors being of only two levels each (type: i/ni,  marker: km/rm), the sphericity hypothesis does not need to be checked. \emph{Post hoc} analyses are done using the Tukey-Kramer procedure.

%\jfp{je crois que la défintion de la variable dépendante, Asubject, n'est pas la bonne. A subject a a été définit avant comme "Asubject is the average pleasantness for each participant". l'ANOVA à deux facteurs à mesures répétées ne porte pas sur la valeur moyenne mais sur la valeur de l'évaluation pour une scene donnée. il faut donc définir une notation appropriée}

@@

The \emph{post hoc} analysis exhibits significative differences between all the groups of observations, notably between the i/km-scenes ($p<0.05$) and the i/rm-scenes ($p<0.01$).

%\jfp{je ne comprends pas les 2 p-value. il ne peut y a voir qu'une seule p-value lorsqu'on compare deux conditions expérimentales}
